SAT Solving for Vesicle Traffic Systems in Cells

### 2017-08-12 paper/bio_MTedit.tex
bio_review_incorporated

In this section, we will describe the model of VTSs and the basic constraints imposed by cell biology.

@@

\textbf{Synthesis:} Given a particular transport graph, a particular labeling of all the compartments and edges, a particular fusion pairing matrix, and a particular regulatory model, we do the following.

\begin{enumerate}
\item We determine which molecules are active on every compartment or vesicle.
\item For every vesicle fusing to a compartment, we determine whether there exists an active pair (one molecule on the vesicle, one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment, and also that every molecular type entering a set of compartments also leaves that set.
%steady state condition.
\end{enumerate}

>>>>

In this section, we will describe the basic constraints imposed by cell biology. These are all incorporated into an abstract model of a VTS, whose properties will then be explored using SMT solvers.

@@

\textbf{Properties of a VTS that satisfies all cell-biological constraints:} Suppose we are given a particular transport graph, a particular labeling of all the compartments and edges, a particular fusion pairing matrix, and a particular regulatory model. This information is then sufficient to check the following properties, which summarise the cell-biological constraints described above:
\begin{enumerate}
\item We can determine which molecules are active on every compartment or vesicle.
\item For every vesicle fusing to a compartment, we can determine whether there exists an active pair (one molecule on the vesicle, one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the compartment, we can verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We can verify that every molecular type entering a compartment also leaves the compartment, and also that every molecular type entering a set of compartments also leaves that set. This is the steady state condition. In the biological literature this is often referred to as “homeostasis” and is a widespread and well-accepted assumption about cellular behaviour, at least over timescales of minutes to hours \cite{mani2017}.
\end{enumerate}

The biological problem often boils down to such an analysis. A cell biologist might determine which molecules flow between which set of compartments, and biochemical experiments could be used to see how these molecules activate one another. We can then ask: is this a complete and consistent description? That is, do all the required properties listed above hold, given what the experimentalists have told us? It is often the case that biological data is missing. For example, only a few of the dozens of molecules involved in real VTSs have been mapped out. Therefore, it is extremely likely that the description provided by the cell biologist is incomplete. We can use our model to find out which properties have failed to hold, and thus prescribe new experiments in order to fill in the missing information.

Can we find a simple test to see whether any information is missing, given the experimental data? We have shown that graph k-connecte furnishes precisely such a test \cite{shukla}. If the data provided by cell-biologists, suitably represented as a graph, does not have a certain degree of connectivity, this implies that some biological data has been missed. (The converse is not true: even if the required connectivity does hold, there might of course be more information missing.)

Our result about k-connectedness being a useful test of missing information \cite{shukla} was obtained using SAT solvers for graphs up to a certain size, and a certain number of molecules. This was due to limitations in how we encoded the problem. Here we present a much more natural encoding that allows our result to be extended to graph sizes and molecule numbers that are typical of those found in real cells.
